Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-07-08 reasoning.tex
edits

Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
top-level judgment uses $I$ for both pre- and post-conditions, because
our focus is on verifying that a \txnimp\ program \emph{preserves}
database's consistency conditions\footnote{The terms \emph{consistency
    condition}, \emph{high-level invariant}, and \emph{integrity
    constraint} all mean the same.}.  A transaction-local RG judgment
does not include a guarantee relation because transaction-local
effects are not visible outside a transaction. Also, the rely relation
($\R$) of the transaction-local judgment is not the same as the
top-level rely relation ($R$) because it must take into account the
transaction's isolation specification ($\I$). Intuitively, $\R$ is $R$
modulo $\I$.  Recall that a transaction writes to its local database
($\stl$), which is then flushed when the transaction commits. Thus,
the guarantee of a transaction depends on the state of its local
database at the commit point. The pre- and post-condition assertions
($P$ and $Q$) in the local judgment facilitate tracking the changes to
the transaction-local state, which eventually helps us prove the
validity of the transaction's guarantee.  Both $P$ and $Q$ are
bi-state assertions; they relate transaction-local database state
($\stl$) to the global database state ($\stg$). Thus, the
transaction-local judgment effectively tracks how transaction-local
and global states change in relation to each other.

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by the interference, i.e., the rely relation $R$. In
conventional RG, stability is defined as follows, where $\sigma$ denotes a
state:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and isolation
specification, we use multiple definitions of stability in
Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
above. We will introduce them as they are encountered.

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step. If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded. Often, this significantly simplifies the form of $\R_l$ irrespective of $R$. For example, when a transaction is executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$ will be the identity function, since if $\stg \neq \stg'$, there does not exist any $\stg_1$ which satisfies the constraint for $\R_l(\stl,\stg,\stg')$.

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $R_c$. $R_c$, like
$R_l$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit captured by the flush
operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of an
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
should not make the interference from $\stg$ to $\stg''$ valid.  This
stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed. Thus, $\I_{ss}$ remains
invalid.

@@

We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in
  the supplementary material.}:

>>>>

Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
top-level judgment uses $I$ for both pre- and post-conditions, because
our focus is on verifying that a \txnimp\ program \emph{preserves} a
databases' consistency conditions\footnote{The terms \emph{consistency
    condition}, \emph{high-level invariant}, and \emph{integrity
    constraint} are used interchangeably throughout the paper.}.  A
transaction-local RG judgment does not include a guarantee relation
because transaction-local effects are not visible outside a
transaction. Also, the rely relation ($\R$) of the transaction-local
judgment is not the same as the top-level rely relation ($R$) because
it must take into account the transaction's isolation specification
($\I$). Intuitively, $\R$ is $R$ modulo $\I$.  Recall that a
transaction writes to its local database ($\stl$), which is then
flushed when the transaction commits. Thus, the guarantee of a
transaction depends on the state of its local database at the commit
point. The pre- and post-condition assertions ($P$ and $Q$) in the
local judgment facilitate tracking the changes to the
transaction-local state, which eventually helps us prove the validity
of the transaction's guarantee.  Both $P$ and $Q$ are bi-state
assertions; they relate transaction-local database state ($\stl$) to
the global database state ($\stg$). Thus, the transaction-local
judgment effectively tracks how transaction-local and global states
change in relation to each other.

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by the interference, i.e., the rely relation $R$. In
conventional RG, stability is defined as follows, where $\sigma$ denotes a
state:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and the
availability of an isolation specification, we use multiple
definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
convey the same intuition as above.

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state
during the last step by a transaction and if the transaction can take
its next step at the current global database state $\stg$, then even
after the interference, the transaction should still be able to take
its next step. If there is no such state $\stg_1$, then the
transaction will be stuck, and the interference can be safely
discarded. Often, this significantly simplifies the form of $\R_l$
irrespective of $R$. For example, when a transaction is executed
against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$
will be the identity function, since if $\stg \neq \stg'$, there does
not exist any $\stg_1$ which satisfies the constraint for
$\R_l(\stl,\stg,\stg')$.

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_l$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of an
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
should not make the interference from $\stg$ to $\stg''$ valid.  This
stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed.

@@

\noindent We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in
  the supplementary material.}:
